---
title: "Data Visualization using ggplot2"
author: "Beesan Alattal"
output:
  html_document: default
  word_document: default
  pdf_document: default
---


```{r}
# Load necessary libraries
library(ggplot2)
library(dplyr)
library(tidyr)
# 1. Read the Data
data(mtcars)
data(iris)

```


```{r}
# pie chart cyl and cars
# Count the number of cars per cylinder category
cyl_data <- mtcars %>%
  count(cyl) %>%
  mutate(perc = round(n / sum(n) * 100, 1))


# Create the Pie Chart
ggplot(cyl_data, aes(x = "", y = n, fill = factor(cyl))) +
  geom_bar(stat = "identity", width = 1) +
  coord_polar(theta = "y") + 
  theme_void() +
  labs(title = "Distribution of Cars by Cylinder Count", fill = "Cylinders") +
  geom_text(aes(label = paste0(perc, "%")), position = position_stack(vjust = 0.5)) +
  scale_fill_manual(values = c("#5F4B8B", "#8A5FBF", "#A67DB8"))
```

```{r}
# bar plot carb and mtcars
# Count the number of cars per carburetor type
carb_data <- count(mtcars, carb)

# Create bar plot with a single purple color
ggplot(carb_data, aes(x = factor(carb), y = n)) +
  geom_bar(stat = "identity", fill = "#5F4B8B", color = "black") +
  scale_y_continuous(breaks = seq(0, max(carb_data$n), by = 1)) +
  labs(title = "Count of Cars by Carburetor Type",
       x = "Carburetor Type",
       y = "Count of Cars") +
  theme_minimal()

```

```{r}
# stacked bar chart presenting 3 different features
# Count the number of cars per gear and cylinder combination
gear_cyl_data <- mtcars %>%
  count(gear, cyl)

# Create stacked bar plot
ggplot(gear_cyl_data, aes(x = factor(gear), y = n, fill = factor(cyl))) +
  geom_bar(stat = "identity") +
  scale_y_continuous(breaks = seq(1, 15, by = 1)) +
  scale_fill_manual(values = c("#5F4B8B", "#8A5FBF", "#A67DB8")) +
  labs(title = "Count of Each Gear Type Segmented by Cylinders",
       x = "Gear Type",
       y = "Count of Cars",
       fill = "Cylinders") +
  theme_minimal()


```

```{r}
# scatter plot to present the correlation 
# Create scatter plot
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_point(color = "#5F4B8B", size = 3, alpha = 0.7) +
  geom_smooth(method = "lm", color = "black", se = FALSE) +
  labs(title = "Scatter Plot of Weight vs MPG",
       x = "Car Weight (1000 lbs)",
       y = "Miles Per Gallon (MPG)") +
  theme_minimal()

```
# here we are repeating the process but different plots choosing the best
```{r}

ggplot(mtcars, aes(x = wt, fill = "Weight")) +
  geom_density(alpha = 0.5, color = "black") +
  labs(title = "Density Plot of Car Weight", x = "Weight (1000 lbs)", y = "Density") +
  theme_minimal() +
  scale_fill_manual(values = "#5F4B8B")

```
```{r}
# Compute correlation matrix
cor_matrix <- cor(mtcars[, c("wt", "mpg")])

# Convert matrix to a tidy format
cor_data <- as.data.frame(cor_matrix) %>%
  pivot_longer(cols = everything(), names_to = "Var2", values_to = "value") %>%
  mutate(Var1 = rep(rownames(cor_matrix), times = ncol(cor_matrix)))

# Create the heatmap
ggplot(cor_data, aes(x = Var1, y = Var2, fill = value)) +
  geom_tile() +
  scale_fill_gradient2(low = "red", mid = "white", high = "blue", midpoint = 0) +
  labs(title = "Heatmap of Correlation Between Variables", x = "", y = "") +
  theme_minimal()

```
# The dot plot is the most suitable, showing clusters and some relations 
```{r}
# Dot Plot is the best choice, because it is showing some relations and the clusters and it does represent valuable insights.
ggplot(mtcars, aes(x = wt, y = mpg)) +
  geom_dotplot(binaxis = 'y', stackdir = 'center', dotsize = 0.5, fill = "#5F4B8B") +
  labs(title = "Dot Plot of Weight vs MPG",
       x = "Weight (1000 lbs)",
       y = "Miles Per Gallon (MPG)") +
  theme_minimal()

```
```{r}
ggplot(mtcars, aes(x = wt)) +
  stat_ecdf(geom = "step", color = "#5F4B8B") +
  labs(title = "ECDF of Car Weight",
       x = "Weight (1000 lbs)",
       y = "Cumulative Probability") +
  theme_minimal()

```
```{r}
ggplot(mtcars, aes(sample = wt)) +
  geom_qq(color = "#5F4B8B") +
  geom_qq_line() +
  labs(title = "Q-Q Plot for Car Weight",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_minimal()

```
# choosing the violin because it is the best of both worlds showing the outliers and the distrubution across the feature 

```{r}

# showing the outliers and the distribution at the same time
# Create a violin plot for Sepal Length across Species
ggplot(iris, aes(x = Species, y = Sepal.Length, fill = Species)) +
  geom_violin(trim = FALSE, alpha = 0.7) +
  scale_fill_manual(values = c("#5F4B8B", "#8A5FBF", "#A67DB8")) +
  labs(title = "Violin Plot of Sepal Length Across Species",
       x = "Species",
       y = "Sepal Length (cm)",
       fill = "Species") +
  theme_minimal()
```



This report presents various visualizations to explore patterns within the `mtcars` and `iris` datasets using `ggplot2`:

- **The pie chart** represents the **proportion of cylinder types** in the `mtcars` dataset.  
- **The bar plot** displays the **count of different carburetor types**.  
- **The stacked bar plot** compares how **gear types are distributed across cylinder categories**.  
- **The scatter plot** highlights the **relationship between car weight (`wt`) and fuel efficiency (`mpg`)**.  
- **The density plot** visualizes the **distribution of car weight**.  
- **The heatmap** provides insights into **correlation intensity between variables**.  
- **The violin plot** effectively demonstrates the **distribution of Sepal Length across different iris species**.




